geochemistry

### 2025-03-23 geochemistry.qmd
Adding chapters and making stuff work
PDF spacing to double in yalm. Solved issues with tinytex

execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
---

@@

# X-ray diffraction (XRD)

@@

xrd_perc <- read_csv('data/xrd_goals.csv')

sp6_xrd <- read_csv('data/SP6_xrd.csv')
sp6_peaks <- read_csv('data/SP6_CSV_PEAKS.csv')

sp6_long <- sp6_xrd %>%
  pivot_longer(cols = c(Difference, Background, Observed, SiO2, Dolomite), 
               names_to = "Phase", 
               values_to = "Count")

p_sp6 <- sp6_long |> 
  mutate(Phase = factor(Phase, levels = c("Difference", "Observed", "SiO2", "Dolomite", "Background"))) |>
  ggplot(aes(x = Baseline, y = Count, color = Phase)) +
  geom_line(size = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity") +
  theme_minimal() +
  geom_point(data = sp6_peaks, aes(x = Angle, y = Intensity, colour = Phase), size = 5,
             shape = 21, show.legend = FALSE) +
  geom_label(data = sp6_peaks, aes(x = Angle, y = Intensity, label = d), 
             size = 3, fill = "white", label.size = 0.2, vjust = -0.8, show.legend = FALSE)

sp18_xrd <- read_csv('data/SP18_xrd.csv')
sp18_peaks <- read_csv('data/SP18_peaks.csv')
sp18_peaks <- sp18_peaks |> 
  dplyr::mutate(Phase = case_when(
    Phase == "SiO2p3221" ~ "SiO2",
    TRUE ~ Phase))
sp18_xrd <- sp18_xrd |> 
  dplyr::mutate(Difference = Observed - Calculated - 3000)

sp18_long <- sp18_xrd %>%
  pivot_longer(cols = c(Difference, Background, Observed, SiO2, Dolomite, Hydroxyapatite), 
               names_to = "Phase", 
               values_to = "Count")

p_sp18 <- sp18_long |> 
  mutate(Phase = factor(Phase, levels = c("Difference" ,"Observed", "SiO2", "Dolomite", "Hydroxyapatite", "Background"))) |>
  ggplot(aes(x = Angle, y = Count, color = Phase)) +
  geom_line(size = 0.5) +
  labs(x = "Diffraction Angle", y = "Intensity") +
  theme_minimal() +
  geom_point(data = sp18_peaks, aes(x = Angle, y = Intensity, colour = Phase), size = 5,
             shape = 21, show.legend = FALSE) +
  geom_label(data = sp18_peaks, aes(x = Angle, y = Intensity, label = d), 
             size = 3, fill = "white", label.size = 0.2, vjust = -0.8, show.legend = FALSE)

@@

# Scanning electron microscopy and energy dispersive X-ray spectroscopy (SEM-EDS)

# Portable X-ray fluorescence (pXRF)

@@

install.packages("ggpubr")
library(ggpubr)

boxplots <- ggarrange(b3, b8, b10, b9, 
          labels = c("a","b","c","d"),
          ncol = 2,
          nrow = 2,
          common.legend = TRUE,
          legend = "bottom")
boxplots_comb2 <- ggarrange(b5, b11, b12, b13, 
          labels = c("a","b","c","d"),
          ncol = 2,
          nrow = 2,
          common.legend = TRUE,
          legend = "bottom")

#ggsave("boxplots_a.png", plot = boxplots_comb1, width = 8, height = 10, dpi = 600, bg = "white")
#ggsave("boxplots_b.png", plot = boxplots_comb2, width = 8, height = 8, dpi = 600, bg = "white")

```{r}
#| label: tbl-occurrences
#| tbl-cap: A table.
#| include: false

combined_table <- pxrf_raw |>
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb") |> 
  mutate(across(
    .cols = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb"),
    .fns = ~ case_when(
      . == 0 | . == "< LOD" ~ "No",
      . > 0 ~ "Yes",
      TRUE ~ as.character(.))
    )) |> 
  mutate(`Chert type` = case_when(
    `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "TYPE 2",
    `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "TYPE 6",
    TRUE ~ `Formation/Type`))

tbl_MgO <- combined_table |> 
  tabyl(`Chert type`, MgO) |>
  adorn_percentages("row") |> 
  adorn_pct_formatting(digits = 1) |> 
  select("Chert type", Yes) |> 
  rename(MgO = Yes)
# Al2O3
tbl_Al2O3 <- combined_table |> 
  tabyl(`Chert type`, Al2O3) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>
  select("Chert type", Yes) |> 
  rename(Al2O3 = Yes)
# SiO2
tbl_SiO2 <- combined_table |> 
  tabyl(`Chert type`, SiO2) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>
  select("Chert type", Yes) |> 
  rename(SiO2 = Yes)
# P2O5
tbl_P2O5 <- combined_table |> 
  tabyl(`Chert type`, P2O5) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(P2O5 = Yes)
# S
tbl_S <- combined_table |> 
  tabyl(`Chert type`, S) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(S = Yes)
# Cl
tbl_Cl <- combined_table |> 
  tabyl(`Chert type`, Cl) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Cl = Yes)
# K2O has 100% rate
tbl_K2O <- combined_table |> 
  tabyl(`Chert type`, K2O) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(K2O = Yes)
# CaO
tbl_CaO <- combined_table |> 
  tabyl(`Chert type`, CaO) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(CaO = Yes)
# TiO2
tbl_TiO2 <- combined_table |> 
  tabyl(`Chert type`, TiO2) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(TiO2 = Yes)
# V
tbl_V <- combined_table |> 
  tabyl(`Chert type`, V) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(V = Yes)
# MnO
tbl_MnO <- combined_table |> 
  tabyl(`Chert type`, MnO) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(MnO = Yes)
# Fe2O3 has 100% rate
tbl_Fe2O3 <- combined_table |> 
  tabyl(`Chert type`, Fe2O3) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Fe2O3 = Yes)
# Cu
tbl_Cu <- combined_table |> 
  tabyl(`Chert type`, Cu) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Cu = Yes)
# As
tbl_As <- combined_table |> 
  tabyl(`Chert type`, As) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(As = Yes)
# Rb
tbl_Rb <- combined_table |> 
  tabyl(`Chert type`, Rb) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Rb = Yes)
# Sr
tbl_Sr <- combined_table |> 
  tabyl(`Chert type`, Sr) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Sr = Yes)
# Y
tbl_Y <- combined_table |> 
  tabyl(`Chert type`, Y) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Y = Yes)
# Zr
tbl_Zr <- combined_table |> 
  tabyl(`Chert type`, Zr) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Zr = Yes)
# Mo
tbl_Mo <- combined_table |> 
  tabyl(`Chert type`, Mo) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Mo = Yes)
# Ag
tbl_Ag <- combined_table |> 
  tabyl(`Chert type`, Ag) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Ag = Yes)
# Ba
tbl_Ba <- combined_table |> 
  tabyl(`Chert type`, Ba) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Ba = Yes)
# Ce
tbl_Ce <- combined_table |> 
  tabyl(`Chert type`, Ce) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Ce = Yes)
# W
tbl_W <- combined_table |> 
  tabyl(`Chert type`, W) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(W = Yes)
# Au
tbl_Au <- combined_table |> 
  tabyl(`Chert type`, Au) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>

  select("Chert type", Yes) |> 
  rename(Au = Yes)
# Pb
tbl_Pb <- combined_table |> 
  tabyl(`Chert type`, Pb) |>
  adorn_percentages("row") |>                   
  adorn_pct_formatting(digits = 1) |>
  select("Chert type", Yes) |> 
  rename(Pb = Yes)

combined_tbl <- tbl_MgO |> 
  full_join(tbl_Al2O3, by = "Chert type") |> 
  full_join(tbl_SiO2, by = "Chert type") |>
  full_join(tbl_P2O5, by = "Chert type") |> 
  full_join(tbl_S, by = "Chert type") |> 
  full_join(tbl_Cl, by = "Chert type") |> 
  full_join(tbl_K2O, by = "Chert type") |> 
  full_join(tbl_CaO, by = "Chert type") |> 
  full_join(tbl_TiO2, by = "Chert type") |> 
  full_join(tbl_V, by = "Chert type") |> 
  full_join(tbl_MnO, by = "Chert type") |> 
  full_join(tbl_Fe2O3, by = "Chert type") |> 
  full_join(tbl_Cu, by = "Chert type") |> 
  full_join(tbl_As, by = "Chert type") |> 
  full_join(tbl_Rb, by = "Chert type") |> 
  full_join(tbl_Sr, by = "Chert type") |> 
  full_join(tbl_Y, by = "Chert type") |> 
  full_join(tbl_Zr, by = "Chert type") |> 
  full_join(tbl_Mo, by = "Chert type") |> 
  full_join(tbl_Ag, by = "Chert type") |> 
  full_join(tbl_Ba, by = "Chert type") |> 
  full_join(tbl_Ce, by = "Chert type") |> 
  full_join(tbl_W, by = "Chert type") |> 
  full_join(tbl_Au, by = "Chert type") |> 
  full_join(tbl_Pb, by = "Chert type")

combined_tbl[, 2:26] <- lapply(combined_tbl[, 2:26], function(x) as.numeric(sub("%", "", x)))
gt(combined_tbl) %>%
  data_color(
    columns = vars("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb"),  # Multiple columns
    colors = scales::col_numeric(
      palette = c("#f8f0fc", "#dab6fc", "#b6f7c1"), # Gradient from red to green
      domain = c(0, 100))
  )

@@

## a) PCA geological regional cherts
pca_db_algarve <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, S, Cl, K2O, CaO, TiO2, Fe2O3) |>
  filter(`Formation/Type` %in% c("LJW", "MJE", "UJE", "UJW", "TW"))
# Convert "< LOD" values to NA
pca_db_algarve[pca_db_algarve == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_algarve$`Formation/Type` <- as.factor(pca_db_algarve$`Formation/Type`)  # Ensure it's a factor
group_var_algarve <- pca_db_algarve$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_alg <- pca_db_algarve %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_alg <- complete.cases(df_numeric_alg)  # Identify complete rows
df_numeric_alg <- df_numeric_alg[complete_cases, ]  # Remove NA rows
group_var_algarve <- group_var_algarve[complete_cases]  # Filter group_var accordingly
# Scale values
df_scaled_alg <- scale(df_numeric_alg)
# Ensure row names are properly set
rownames(df_scaled_alg) <- make.names(group_var_algarve, unique = TRUE)
# Perform PCA
pca_result_alg <- PCA(df_scaled_alg, graph = FALSE)
# Summary of PCA
summary(pca_result_alg)
fviz_eig(pca_result_alg, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_alg <- fviz_pca_biplot(pca_result_alg, 
                label = "var", 
                habillage = group_var_algarve,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.95) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_alg <- data.frame(pca_result_alg$ind$coord, Group = group_var_algarve)
# PCA plot with groups
scatter_alg <- ggplot(df_pca_alg, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

>>>>

execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600

---

@@

## X-ray diffraction (XRD)

Some text.
Some other text.
A bit more of text.

@@

@@

## Scanning electron microscopy and energy dispersive X-ray spectroscopy (SEM-EDS)

## Portable X-ray fluorescence (pXRF)

@@

library(ggpubr)

boxplots <- ggarrange(b3, b8, b10, b9, 
          labels = c("a","b","c","d"),
          ncol = 2,
          nrow = 2,
          common.legend = TRUE,
          legend = "bottom")

@@

## a) PCA geological regional cherts
pca_db_algarve <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, S, Cl, K2O, CaO, TiO2, Fe2O3) |>
  filter(`Formation/Type` %in% c("LJW", "MJE", "UJE", "UJW", "TW"))
# Convert "< LOD" values to NA
pca_db_algarve[pca_db_algarve == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_algarve$`Formation/Type` <- as.factor(pca_db_algarve$`Formation/Type`)  # Ensure it's a factor
group_var_algarve <- pca_db_algarve$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_alg <- pca_db_algarve %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_alg <- complete.cases(df_numeric_alg)  # Identify complete rows
df_numeric_alg <- df_numeric_alg[complete_cases_alg, ]  # Remove NA rows
group_var_algarve <- group_var_algarve[complete_cases_alg]  # Filter group_var accordingly
# Scale values
df_scaled_alg <- scale(df_numeric_alg)
# Ensure row names are properly set
rownames(df_scaled_alg) <- make.names(group_var_algarve, unique = TRUE)
# Perform PCA
pca_result_alg <- PCA(df_scaled_alg, graph = FALSE)
# Summary of PCA
summary(pca_result_alg)
fviz_eig(pca_result_alg, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_alg <- fviz_pca_biplot(pca_result_alg, 
                label = "var", 
                habillage = group_var_algarve,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.95) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_alg <- data.frame(pca_result_alg$ind$coord, Group = group_var_algarve)
# PCA plot with groups
scatter_alg <- ggplot(df_pca_alg, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")
